%!TEX root = ../dissertation.tex
\chapter{Conclusion}\label{chapt:conclusion}

% =====================================================================
%  DRAFT v1 (Asma) -- full prose. ~6 pages target. No NEW results here.
% =====================================================================

\section{Summary}\label{sect:concl_summary}

This thesis studied zero-shot, instance-level composed image retrieval on the
i-CIR benchmark~\cite{psomas2025icir}, where a query is a reference image
together with a short modification text and the goal is to retrieve the specific
\emph{instance} the user has in mind, modified as requested. The core
contribution is to add a third retrieval modality on top of the usual image and
text signals: a frozen multimodal large language model (InternVL-3.5-8B) reads
the reference image and the modification together and writes a short description
of the \emph{imagined target}, and the embedding of this \emph{target caption}
is fused with the image and text similarities by a clamped triplet product. The
entire pipeline is zero-shot -- every encoder, the caption generator, and all
fusion and post-processing steps are off-the-shelf and frozen. We developed and
ablated each component on a $20\%$ development subset and then re-ran the
successful configurations at full scale, reporting all headline numbers on the
full $752$K-image database with four backbones.

\section{Key Findings}\label{sect:concl_findings}

The experiments support four conclusions.
\begin{itemize}
  \item \textbf{The caption is the single most effective component.} Adding the
        target caption improves every backbone on the full dataset, and in the
        component ablation it is the largest single contributor -- larger than
        feature centering and far larger than the remaining \textsc{Basic}
        preprocessing.
  \item \textbf{A simple caption can replace elaborate preprocessing.} A plain
        text$\times$image baseline plus the caption reaches the level of the
        fully tuned \textsc{Basic} pipeline, and with a strong backbone exceeds
        it. The textual signal is better supplied by an explicit target
        description than by contextualizing the bare modification phrase.
  \item \textbf{Centering and the caption are complementary.} Their gains add
        up; the best configuration combines both.
  \item \textbf{SigLIP2 is the strongest backbone for the pipeline,} reaching
        the thesis-best $61.98$ macro-mAP with the caption, centering, and
        contextualisation, even though CLIP-H is stronger on the bare duplet
        baseline -- the backbone must be judged within the full pipeline.
\end{itemize}

\section{Limitations}\label{sect:concl_limitations}

Several limitations qualify these findings. The approach is \emph{zero-shot by
design}, so it leaves on the table any gains a learned fusion could provide.
Caption generation introduces \emph{variance}: a single instruction is
prompt-sensitive, which the \texttt{avg-5} averaging mitigates but does not
eliminate, and it adds a non-trivial \emph{compute cost} (five MLLM generations
per query). The ablations were conducted on a $20\%$ \emph{development subset};
although the subset baseline closely matches the full-dataset baseline, not
every ablated configuration was re-run at full scale. Finally, the caption
relies on a \emph{single MLLM} (InternVL-3.5-8B), and the sensitivity of the
results to the choice of captioner was not explored.

\section{Future Work}\label{sect:concl_future}

Several directions follow naturally. A stronger or more efficient
\emph{captioner} -- or a single high-quality caption that matches \texttt{avg-5}
without five generations -- would cut the main compute cost. The fixed,
unweighted product could be replaced by a \emph{learned or weighted fusion} of
the three similarities, which would relax the zero-shot constraint in exchange
for accuracy. Because recall is already high -- the correct instance lands in the
top-ranked shortlist for the large majority of queries (Recall@10 reaches
$92.2\%$ on the best backbone) -- a \emph{second-stage reranker} that reorders
this shortlist is a promising direction: most of the remaining error is in the
\emph{ordering} of candidates that are already retrieved. A
\emph{text-conditioned} reranker that reorders by similarity to the target
caption, rather than to the reference image, is a natural candidate, since it
keeps the modification in the loop at the reranking stage. Finally, the
caption-as-third-modality idea is benchmark-agnostic and could be evaluated
\emph{beyond i-CIR}, on category-level CIR benchmarks and other instance-level
retrieval settings, to test how far the explicit target-description signal
generalises.
