%!TEX root = ../dissertation.tex
\chapter{Related Work}\label{chapt:related}

This chapter situates the proposed method within the composed image retrieval
literature. Section~\ref{sect:rel_benchmarks} traces the evolution of CIR
benchmarks from semantic-level datasets to the instance-level evaluation
protocol, and examines the unimodal shortcuts that pervade existing benchmarks.
Section~\ref{sect:rel_zeroshot} surveys the main families of zero-shot and
training-free methods. Section~\ref{sect:rel_discussion} discusses cross-cutting
open challenges, and Section~\ref{sect:rel_gap} identifies the specific gap this
thesis addresses.

% =====================================================================
\section{Benchmarks and Their Limitations}\label{sect:rel_benchmarks}
% =====================================================================

\subsection{Semantic-Level Datasets}\label{sect:rel_semantic}

The composed image retrieval task was popularised by TIRG (Text Image Residual
Gating)~\cite{vo2019tirg}, which proposed an \emph{early-fusion} network that
combines reference-image and modification-text features through a gated residual
connection, trained with a metric-learning loss on annotated triplets. TIRG
established the benchmark-plus-supervised-fusion recipe that defined the field
for several years.

The most widely used benchmarks that followed are evaluated at a
\emph{semantic} level: a retrieved image is correct if it belongs to the same
broad category as the intended target.
Fashion-IQ~\cite{wu2021fashioniq} targets fashion product retrieval with
natural-language relative feedback; CIRR~\cite{liu2021cirr} moves to
open-domain images and highlights fine-grained visual differences;
CIRCO~\cite{baldrati2023circo} provides multiple validated ground-truth targets
per query to reduce annotation noise. A recurring limitation of all three is
that they are constructed by automatically pairing visually similar images and
then generating or crowdsourcing a description of their difference -- a process
that produces ambiguous or redundant modification texts~\cite{psomas2025icir}.
Because relevance is defined at the category level, evaluation pools contain few
challenging negatives and exhibit high false-negative rates, which inflate the
apparent performance of CIR systems.

\subsection{Unimodal Shortcuts: the CIRCUS Audit}\label{sect:rel_circus}

The severity of these problems is quantified by CIRCUS~\cite{attimonelli2026circus},
a systematic audit of four widely used CIR benchmarks (CIRR, Fashion-IQ, LaSCo,
and CIRCO) evaluated across eleven generalist multimodal embedding models. The
study finds that between 32.2\,\% and 83.6\,\% of queries can be solved using
only the text or only the image; high benchmark performance can therefore arise
from unimodal signals rather than genuine multimodal composition.
Human validation of 4\,741 shortcut-free queries reveals that only 1\,689
(35.6\,\%) are well-formed: the remainder contain ambiguous edits or mismatched
targets. Performance drops sharply on this cleaned subset -- the Recall on CIRR
falls from 70.1 to 24.1 for a state-of-the-art model -- confirming that much of
the apparent progress in CIR is attributable to exploiting shortcuts rather than
learning to compose modalities.

\subsection{Instance-Level CIR: the i-CIR Benchmark}\label{sect:rel_icir}

To address the limitations of semantic-level evaluation, i-CIR~\cite{psomas2025icir}
defines correctness at the level of a specific object \emph{instance}: a
retrieved image is correct only if it depicts the \emph{same instance} as the
reference image, modified as the query text requests. The benchmark is
deliberately engineered around three types of hard negatives -- \emph{visual}
negatives (the same or similar instance under a different modification),
\emph{textual} negatives (images that satisfy the modification text but show a
different instance), and \emph{composed} negatives (the correct instance under a
different modification) -- so that a model which exploits only one modality is
systematically penalised. The detailed dataset statistics and evaluation protocol
are described in Chapter~\ref{chapt:method}.

Alongside the benchmark, i-CIR contributes \textsc{Basic}, a training-free
zero-shot baseline that combines feature centering, dimensionality reduction,
query expansion, and contextualization of the modification text into a
late-fusion multiplicative score. \textsc{Basic} demonstrates that careful use
of frozen features alone is a deceptively strong baseline, and it is the method
this thesis extends (Chapter~\ref{chapt:method}).

% =====================================================================
\section{Zero-Shot and Training-Free Methods}\label{sect:rel_zeroshot}
% =====================================================================

Because supervised fusion requires expensive triplet annotation and generalises
poorly across domains, a substantial line of work removes CIR-specific training
entirely. We group these methods into four families.

\subsection{Textual Inversion}\label{sect:rel_textual_inv}

Textual inversion methods avoid triplet supervision by learning a mapping --
using only image--caption data -- that converts a reference image into a
\emph{pseudo-word token}. The modification text is then combined with this token
in a natural-language prompt, which the frozen text encoder of a
vision--language model embeds as a composed query.
Pic2Word~\cite{saito2023pic2word} introduced this idea for CIR, showing that a
lightweight mapping network trained on image--caption pairs enables zero-shot
retrieval without any annotated triplets. SEARLE~\cite{baldrati2023circo} refines
the inversion by optimising the pseudo-token to be close to a set of descriptive
words, and it is released alongside the CIRCO benchmark as a more challenging
evaluation. LinCIR~\cite{gu2024lincir} takes the idea further by training the
mapping using language only, without any images.

These methods share the philosophy of routing the query through the text branch
so that a frozen text encoder can handle it -- the same philosophy this thesis
adopts -- but they compress the reference into a single learned token. That
single slot must carry all relevant visual context, which limits expressiveness
for novel instances or fine-grained modifications. By contrast, this thesis uses
a frozen MLLM to produce a full, human-readable description of the imagined
target, making the composed signal explicit rather than opaque.

\subsection{Self-Supervised and Diffusion-Based Methods}\label{sect:rel_generative}

\paragraph{MagicLens.}
MagicLens~\cite{zhang2024magiclens} extends CIR to open-ended instructions by
exploiting image pairs that naturally co-occur on the same web page. Such pairs
share implicit semantic relations (e.g.\ ``inside view of'' or ``charger of a
product''); foundation models (PaLI and PaLM\,2) are used to synthesise
instructions describing these relations, yielding a dataset of 36.7 million
triplets with diverse, naturally occurring semantics. Trained on this corpus,
MagicLens achieves state-of-the-art performance across eight retrieval
benchmarks and handles search intents that go beyond visual similarity.
Its success demonstrates that large-scale self-supervision can provide
diverse compositional instructions without any manual triplet annotation.

\paragraph{CompoDiff.}
CompoDiff~\cite{gu2024compodiff} frames CIR as a conditional latent-diffusion
problem: a diffusion model learns to translate the embedding of the reference
image into the embedding of the target, guided by the text condition. To supply
training data, the authors construct SynthTriplets18M, a dataset of 18.8
million automatically generated triplets, which supports versatile conditions
including negative text and image masks. The diffusion formulation offers
controllability -- the relative influence of the image and text can be adjusted
at inference time -- but it requires running a generative process for every
query and training on a large synthetic dataset, which may introduce a domain
gap relative to real images.

\subsection{Vision-by-Language and MLLM Reasoning}\label{sect:rel_caption}

\paragraph{CIReVL.}
CIReVL~\cite{karthik2024cirevl} pioneered the \emph{vision-by-language}
paradigm for CIR: it uses a generative VLM (e.g.\ BLIP-2 or CoCa) to caption
the reference image, prompts a separate LLM to rewrite that caption according to
the modification text, and retrieves images using a frozen CLIP encoder applied
to the rewritten caption. The pipeline is entirely training-free and modular, and
its intermediate representation -- a natural-language description of the intended
target -- is human interpretable. CIReVL confirms that an explicit textual
description of the target is a powerful zero-shot query; however, the sequential
two-stage design means the captioner has no access to the modification text, and
the LLM has no direct access to the image, so information can be lost at each
handoff.

\paragraph{MCoT-RE.}
MCoT-RE~\cite{park2025mcot} addresses the information-loss problem of sequential
pipelines by guiding an MLLM through a \emph{multi-faceted chain-of-thought}
(CoT) process that jointly attends to the modification instruction and the
broader visual context. The MLLM generates two distinct captions: a
\emph{modification-focused} caption describing the explicitly stated change, and
an \emph{integration-focused} caption that also incorporates contextual visual
cues. Candidates are first filtered using the modification-focused caption; a
multi-grained re-ranking step then uses both captions together with the reference
image to produce the final ranking. MCoT-RE achieves state-of-the-art results
among training-free methods, improving Recall@10 on FashionIQ by 6.24\,\% and
Recall@1 on CIRR by 8.58\,\% over prior work. Among all surveyed methods,
MCoT-RE is the most closely related to this thesis: both use a frozen MLLM to
generate caption-based retrieval signals rather than a pseudo-token. The key
differences are that this thesis feeds the reference image and modification text
\emph{jointly} to the MLLM in a single step, uses the generated caption as a
\emph{primary retrieval modality} fused at query time rather than for reranking,
and is evaluated on the instance-level i-CIR benchmark rather than on
category-level datasets.

\paragraph{MCL.}
MCL (Multimodal Composition Learning)~\cite{li2024mcl} takes a training-based
approach to improving cross-modal context understanding in MLLMs. Rather than
translating one modality into the other, MCL introduces two tasks --
Multimodal-Context Captioning (MC-Cap) and Multimodal-Context Retrieval
(MC-Ret) -- that require the model to generate or retrieve information
conditioned on both modalities simultaneously. A MultiModal Composition dataset
of 2.7 million tuples is constructed to supply supervision. The resulting model
performs zero-shot CIR as well as visual storytelling and visual dialogue,
demonstrating that improved cross-modal composition generalises broadly.

\subsection{Domain Conversion}\label{sect:rel_domain}

FREEDOM~\cite{efthymiadis2025freedom} addresses image-domain conversion --
retrieving the content of a query image in a new domain specified by text --
without any task-specific training. Rather than mapping the reference image to a
continuous pseudo-word token, FREEDOM performs nearest-neighbour search in a
discrete word vocabulary and treats the result as a soft distribution over
words; the mapped words are combined with the domain text as a weighted ensemble
query embedded by CLIP's text encoder. Evaluated on three new domain-conversion
benchmarks, FREEDOM outperforms continuous pseudo-word methods and confirms that
a frozen VLM contains sufficient knowledge for open-domain retrieval when paired
with simple, principled post-processing.

For remote-sensing retrieval, WEICOM~\cite{psomas2024rscir} introduces a CIR
benchmark covering modifications such as colour, context, and density, and
balances image and text similarities via a scalar weighting parameter
$\lambda \in [0,1]$; it appears as a comparison point in
Chapter~\ref{chapt:results}.

% =====================================================================
\section{Discussion}\label{sect:rel_discussion}
% =====================================================================

\subsection{Dataset Quality and Unimodal Shortcuts}\label{sect:rel_disc_datasets}

A central lesson from the CIRCUS audit~\cite{attimonelli2026circus} is that
apparent progress in CIR does not necessarily reflect genuine multimodal
composition. When datasets are constructed from automatic image similarity
search and modifications that correlate with category-level cues, high retrieval
scores are achievable without any cross-modal fusion. Human validation further
reveals that many queries in existing datasets are ill-formed, making benchmark
improvement unreliable. These findings motivate instance-level evaluation with
deliberately designed hard negatives of each type, as i-CIR~\cite{psomas2025icir}
provides: a method cannot succeed by latching onto either modality alone because
each modality's hard negatives are present in the pool.

\subsection{Joint Multimodal Processing and Modality Balance}\label{sect:rel_disc_joint}

A recurring limitation of sequential pipelines is the loss of cross-modal
context at each handoff. When a captioner describes the reference image without
knowing the modification, and an LLM edits that caption without seeing the
image, neither step has the full picture~\cite{karthik2024cirevl}. MCoT-RE~\cite{park2025mcot}
mitigates this by using a single MLLM that attends to both modalities; this
thesis adopts the same principle. The \textsc{Basic}~\cite{psomas2025icir}
late-fusion approach sidesteps the issue differently by keeping image and text
signals separate and combining their scores multiplicatively, but at the cost of
a weak textual signal: a short, contextualised modification phrase is a poor
standalone text query compared to a full description of the intended target.

A related challenge is modality balance. WEICOM's scalar
$\lambda$~\cite{psomas2024rscir} and CompoDiff's~\cite{gu2024compodiff}
controllable conditioning both allow the relative weight of image versus text to
be tuned. This thesis instead adopts a fixed clamped triplet product
$\clamp(\sv)\cdot\clamp(\st)\cdot\clamp(\sc)$, which imposes a soft logical
\textsc{and}: a candidate must agree with the reference image, the modification
text, \emph{and} the generated caption to score highly, without requiring any
manual tuning of weights.

Computational cost is also a practical consideration. Diffusion-based
methods~\cite{gu2024compodiff} and MLLM reasoning pipelines run heavy models at
inference time; \textsc{Basic}~\cite{psomas2025icir} and
FREEDOM~\cite{efthymiadis2025freedom} represent the lightweight extreme of pure
post-processing. The caption generation step of this thesis is heavier than
pure post-processing but is run once per query offline, after which retrieval
reduces to the same inner-product computation as \textsc{Basic}.

% =====================================================================
\section{Positioning and Gap}\label{sect:rel_gap}
% =====================================================================

In summary, the literature presents a spectrum from supervised fusion methods
(expensive annotation, poor transfer) through textual-inversion methods (no
triplets needed, but the reference is compressed into a single opaque token)
to training-free, language-mediated methods (the reference is described in full
text). The strongest training-free baseline on the instance-level benchmark,
\textsc{Basic}~\cite{psomas2025icir}, contextualises the bare modification phrase
rather than describing the intended target, leaving the textual signal weak for
short or fragmentary queries. The closest competing method,
MCoT-RE~\cite{park2025mcot}, also uses an MLLM to generate caption-based signals,
but is evaluated only on category-level benchmarks and uses the captions for
candidate filtering and reranking rather than as a primary retrieval modality
fused at query time; it provides no evidence that the approach generalises to
instance-level retrieval, where both visual identity and precise modification
must be respected simultaneously.

The gap this thesis addresses is therefore: \emph{in the zero-shot, instance-level
setting, can a frozen MLLM that jointly reads the reference image and modification
text generate a target caption that, when fused as a third retrieval modality
alongside the image and text signals, provides a stronger and simpler textual
signal than the elaborate contextualisation of \textsc{Basic}}? The remainder of
the thesis answers this question affirmatively, and the ablation studies in
Chapter~\ref{chapt:results} quantify the contribution of each component of the
proposed pipeline.
