%!TEX root = ../dissertation.tex

Composed image retrieval searches an image database with a query that pairs a
reference image with a short modification text specifying how the desired result
should differ from it. The task is challenging: a system must interpret the
reference image, apply the requested modification, and pick out the matching
items among many near-identical distractors. The recent i-CIR benchmark
intensifies this difficulty by defining relevance at the level of a specific
object \emph{instance} rather than a general category, and by surrounding every
query with hard negatives that penalise any method relying on a single modality.
Its accompanying baseline, \textsc{Basic}, combines frozen vision--language
features with careful, training-free post-processing and proves surprisingly
competitive. Its weakness, however, is that it reads the modification text on its
own, without the reference image: an instruction such as ``make it red'' tells a
text encoder almost nothing, because the encoder cannot see \emph{what} should
become red.

The idea of this thesis is to let language carry the composition. A frozen
multimodal large language model reads the reference image and the modification
instruction \emph{together} and writes a short description of the \emph{imagined
target}, performing the requested edit in words rather than in pixels. This
self-contained \emph{target caption} is embedded by the frozen text encoder and
fused with the image and text similarities through a clamped triplet product that
behaves as a soft logical \textsc{and}: a candidate must agree with all three
signals to rank highly. The whole pipeline is \emph{zero-shot} -- every encoder,
the caption generator, and all fusion and post-processing steps are off-the-shelf
and frozen. We choose this training-free setting deliberately: supervised CIR
methods depend on large collections of (reference image, modification text,
target image) triplets that are hard to obtain at scale and, however they are
collected, cover far less visual and textual variety than the data behind modern
vision--language pre-training; the reasons are developed in the introduction.

The method is developed on an instance subset of i-CIR, and the successful
configurations are re-run at full scale across several vision--language
backbones. The target caption is consistently the single most effective
component: it improves every backbone, and on its own, added to a plain
image--text baseline, it is already competitive with the entire \textsc{Basic}
post-processing stack it is designed to replace. On the standard CLIP ViT-L/14
backbone of the original benchmark, the proposed method reaches $35.76$
macro-mAP, surpassing both the $34.35$ reported for the published \textsc{Basic}
pipeline and our own reproduction of it ($32.48$ in an identical software
environment); the best overall configuration attains $61.98$ macro-mAP, the
strongest result reported in this thesis.
