%!TEX root = ../dissertation.tex
I am deeply grateful to my supervisor, Professor Lamberto Ballan, and to my
co-supervisor, Marco Fiorucci, for their guidance, insight, and patience
throughout this work. Their feedback shaped both the direction of this thesis
and the way I have learned to think about research.

I would also like to thank the Department of Mathematics ``Tullio Levi-Civita''
and its faculty for the education and the environment that made this work
possible. I am especially thankful for access to the department's HPC cluster and
GPU resources: this thesis would not have been feasible without them, and I am
grateful to the department for allowing me to use and experiment with this
infrastructure. My sincere thanks go to Luca Righi, who is responsible for the
cluster, for his help and availability whenever I ran into a problem --- his
support made a real difference at many points along the way.

Finally, and most importantly, I want to thank my family and the family I chose,
my friends. Thank you for your unconditional love and your constant support,
through every difficulty and every long day. This achievement is as much yours as
it is mine, and I could not have reached it without you.

\vspace{2\baselineskip}
\begin{center}
\itshape
This thesis is dedicated to all the Afghan and Iranian girls who were denied the
chance to pursue their education --- not for any lack of will or talent, but
because of oppression and the accident of where they happened to be born. This
work is for them.
\end{center}
